Simulating High-Dimensional Multivariate Data using the \pkg{bigsimr} Package

### 2021-06-01 article_bigsimr.tex
Revise for JSS submission pre cross ref fix

This section describes a low-dimensional (2D) random vector simulation
workflow via the \texttt{bigsimr} \texttt{R} package
(\url{https://github.com/SchisslerGroup/r-bigsimr}). The package
\texttt{bigsimr} provides an interface to the native code written in
Julia, registered as the \texttt{Bigsimr} Julia package
(\url{https://github.com/adknudson/Bigsimr.jl}). In addition to the
native Julia \texttt{Bigsimr} package and \texttt{R} interface
\texttt{bigsimr}, we also provide a python interface \texttt{bigsimr}
(\url{https://github.com/SchisslerGroup/python-bigsimr/}) that
interfaces with the Julia \texttt{Bigsimr} package. The Julia package
provides a high-performance implementation of our proposed random vector
generation algorithm and associated functions (see Section
@ref(algorithms)).

@@

\hypertarget{bivariate-example-description}{%
\subsection{Bivariate example
description}\label{bivariate-example-description}}

We illustrate the use of \texttt{bigsimr} motivated by the New York air
quality data set (\texttt{airquality}) included in the R
\texttt{datasets} package. First, we load the \texttt{bigsimr} library
and a few other convenient data science packages, including the
syntactically-elegant \texttt{tidyverse} suite of \texttt{R} packages.
The code chunk below prepares the computing environment:

@@

For simplicity and to provide a minimal working example, we consider
bivariate simulation of the two \texttt{airquality} variables
\texttt{Temperature}, in degrees Fahrenheit, and \texttt{Ozone} level,
in parts per billion.

@@

Figure @ref(fig:ch030-aq-joint-dist) visualizes the bivariate
relationship between Ozone and Temperature. We aim to simulate random
two-component vectors mimicking this structure. The margins are not
normally distributed; particularly the Ozone level exhibits a strong
positive skew.

@@

\caption[Bivariate scatterplot of Ozone vs]{Bivariate scatterplot of Ozone vs. Temp with estimated marginal densities. We model the Ozone data as marginally log-normal and the Temperature data as normal.}\label{fig:ch030-aq-joint-dist}
\end{figure}
\end{CodeChunk}

Next, we specify the marginal distributions and correlation coefficient
(both type and magnitude). Here the analyst is free to be creative. For
this example, we avoid goodness-of-fit considerations to determine the
marginal distributions. But it seems sensible without domain knowledge
to estimate these quantities from the data, and \texttt{bigsimr}
contains fast functions designed for this task.

@@

Based on the estimated densities in Figure
@ref(fig:ch030-aq-joint-dist), we assume \texttt{Temp} is normally
distributed and \texttt{Ozone} is log-normally distributed, as the
latter values are positive and skewed. We use the well-known, unbiased
estimators for the normal distribution's parameters and maximum
likelihood estimators for the log-normal parameters:

@@

Next, we configure the input marginals for later input into
\texttt{rvec}. The marginal distributions are specified using
\texttt{Julia}'s \texttt{Distributions} package and stored in a vector.

@@

As mentioned, the user must decide how to describe correlation based on
the particulars of the problem. For non-normal data and for improved
simulation accuracy/scalability in our scheme, we advocate the use of
Spearman's \(\rho\) correlation matrix \(R_S\) and Kendall's \(\tau\)
correlation matrix \(R_K\). We also support Pearson correlation
coefficient matching, while cautioning the user to check the performance
for their parametric multivariate model (see
\protect\hyperlink{simulations}{Monte Carlo evaluations} below for
evaluation strategies and guidance). These estimation methods are
classical approaches, not designed for high-dimensional correlation
estimation (see the \href{\%7B\#discussion}{Conclusion and Discussion}
sections for more on this).

@@

First we use \texttt{cor\_bounds} to ensure that the pairwise target
correlation values are valid prior to the mapping step which constructs
\(R_X\) for the MVN input into NORTA (see Section @ref(algorithms)).
\texttt{cor\_bounds} estimates the pairwise lower and upper correlation
bounds using the Generate, Sort, and Correlate algorithm of
\citet{DH2011}.

\begin{CodeChunk}
\begin{CodeInput}
R> bounds <- bs$cor_bounds(margins)
R> bounds$lower
\end{CodeInput}
\begin{CodeOutput}
        [,1]    [,2]
[1,]  1.0000 -0.8815
[2,] -0.8815  1.0000
\end{CodeOutput}
\begin{CodeInput}
R> bounds$upper
\end{CodeInput}
\begin{CodeOutput}
       [,1]   [,2]
[1,] 1.0000 0.8803
[2,] 0.8803 1.0000
\end{CodeOutput}
\end{CodeChunk}

Since our single estimated Spearman correlation is within the
theoretical bounds, the correlation is valid as input to \texttt{rvec}.
But even if the 2-dimensional bounds are satisfied for each pair of
margins, this does not guarantee the feasibility of a \(d-\)variate
distribution \citep{BF17}.

To provide higher dimensional feasibility/admissibility checking, we
begin by mapping using \texttt{cor\_convert} and check admissibility.

\begin{CodeChunk}
\begin{CodeInput}
R> # Step 1. Mapping
R> (R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
\end{CodeInput}
\begin{CodeOutput}
       [,1]   [,2]
[1,] 1.0000 0.7886
[2,] 0.7886 1.0000
\end{CodeOutput}
\begin{CodeInput}
R> # Step 2. Check admissibility
R> bs$iscorrelation(R_X)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

The bounds on the Pearson correlation coefficient \(R_P\) between these
margins are restricted as seen below in our MC estimated correlation
bounds:

@@

Our MC estimate of the bounds slightly overestimates the theoretic
bounds of \((-0.881, 0.881)\). An analytic derivation is presented as an
Appendix.

@@

Finally, we arrive at the main function of \texttt{bigsimr}:
\texttt{rvec}. We now simulate \(B=10,000\) random vectors from the
assumed joint distribution of Ozone levels and Temp.

@@

Before applying our methodology to real data simulation, we conduct
several Monte Carlo studies to investigate method performance. Since
marginal parameter matching in our scheme is essentially a sequence of
univariate inverse probability transforms, the challenging aspects are
the accuracy of dependency matching and computational efficiency at high
dimensions. To evaluate our methods in those respects, we design the
following numerical experiments to first assess accuracy of matching
dependency parameters in bivariate simulations and then time the
procedure in increasingly large dimension \(d\).

@@

We select bivariate simulation configurations to ultimately simulate our
motivating discrete-valued RNA-seq example, and, so we proceed in
increasing complexity, leading to the model in our motivating
application in Section @ref(examples). We begin with empirically
evaluating the dependency matching across all three supported
correlations --- Pearson's, Spearman's, and Kendall's --- in identical,
bivariate marginal configurations. For each pair of identical margins,
we vary the target correlation across \(\Omega\), the set of possible
admissible values for correlation type, to evaluate the simulation's
ability to obtain the theoretic bounds. The simulations progress from
bivariate normal, to bivariate gamma (non-normal yet continuous), and
bivariate negative binomial (mimicking RNA-seq counts).

Table @ref(tab:sims) lists our identical-marginal, bivariate simulation
configurations. We increase the simulate replicates \(B\) to check that
our results converge to the target correlations and gauge statistical
efficiency. We select distributions beginning with a standard
multivariate normal (MVN) as we expect the performance to be exact (up
to MC error) for all correlation types. Then, we select a non-symmetric
continuous distribution: a standard (rate =1) two-component multivariate
gamma (MVG). Finally, we select distributions and marginal parameter
values that are motivated by our RNA-seq data, namely values proximal to
probabilities and sizes estimated from the data (see
\href{examples}{Example applications} for estimation details). Thus we
arrive at a multivariate negative binomial (MVNB)
\(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).

Table: (\#tab:sims) Identical margin, bivariate simulation
configurations to evaluate correlation matching accuracy and efficiency.

\begin{table}[]
\begin{tabular}{@{}lcr@{}}
\toprule
Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
$1,000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
\end{tabular}
\end{table}

For each of the unique 9 simulation configurations described above, we
estimate the correlation bounds and vary the correlations along a
sequence of 100 points evenly placed within the bounds, aiming to
explore \(\Omega\). Specifically, we set correlations
\(\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}\),
with \(\hat{l}\) and \(\hat{u}\) being the estimated lower and upper
bounds, respectively, and increment value \(\delta\). The adjustment
factor, \(\epsilon=0.01\), is introduced to handle numeric issues when
the bound is specified exactly.

@@

Overall, the studies show that our methodology is generally accurate
across the entire range of possible correlation values all three
dependency measures, at least in these limited simulation settings for
the rank-based correlations. Our Pearson matching performs nearly as
well as Spearman or Kendall, except for a slight increase in error for
negative binomial case. This is due the particularly large counts
generated with our choice of parameters for \(p\) and \(r\).

@@

With information of our method's accuracy from a low-dimensional
perspective, we now turn to assessing whether \texttt{bigsimr} can scale
to larger dimensional problems with practical computation times.
Specifically, we ultimately generate \(B=1,000\) random vectors for
\(d=\{100, 250, 500, 1000, 2500, 5000, 10000\}\) for each correlation
type, \(\{Pearson, Spearman, Kendall\}\) while timing the algorithm's
major steps.

To mimic the workflow, we first produce a synthetic ``data set'' by
completing the following steps:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Produce heterogeneous gamma marginals by randomly selecting the
  \(j^{th}\) gamma shape parameter from
  \(U_j \sim uniform(1,10), j=1,\ldots,d\) and the \(j^{th}\) rate
  parameter from \(V_j \sim exp(1/5), j=1,\ldots,d\), with the constant
  parameters determined arbitrarily.
\item
  Produce a random full-rank Pearson correlation matrix via
  \texttt{cor\_randPD} of size \(d \times d\).
\item
  Simulate a ``data set'' of \(1,000 \times d\) random vectors via
  \texttt{rvec} (without matching the Pearson exactly). \setstretch{2.0}
\end{enumerate}

With the synthetic data set in hand, we complete and time the following
4 steps:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \begin{enumerate}
  \def\labelenumii{\alph{enumii})}
  \tightlist
  \item
    Estimate the correlation matrix from the ``data'' in the
    \emph{Compute Correlation} step.
  \end{enumerate}
\item
  \begin{enumerate}
  \def\labelenumii{\alph{enumii})}
  \setcounter{enumii}{1}
  \tightlist
  \item
    Map the correlations to initialize the algorithm (Pearson to
    Pearson, Spearman to Pearson, or Kendall to Pearson) in the
    \emph{Adjust Correlation} step.
  \end{enumerate}
\item
  Check whether the mapping produces a valid correlation matrix and, if
  not, find the nearest PD correlation matrix in the \emph{Check
  Admissibility} step.
\item
  Simulate \(1,000\) vectors via the Inverse Transform method in the
  \emph{Simulate Data} step. \setstretch{2.0}
\end{enumerate}

The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core
Intel Core i9 processor, with all 16 threads employed during
computation. The results for \(d \leq 500\) are fast with all times
(except Kendall) under 3 seconds. Table @ref(tab:ch040-moderateDtab)
displays the total computation time of the algorithms steps 1 through 3,
including estimation step i.

@@

\caption{\label{tab:tab:ch040-moderateDtab}Total time to produce 10,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
\centering
\begin{tabular}[t]{llr}
\toprule
Dimension & Correlation type & Total Time (Seconds)\\
\midrule
100 & Pearson & 0.081\\
100 & Spearman & 0.027\\
100 & Kendall & 0.075\\
\addlinespace
250 & Pearson & 0.378\\
250 & Spearman & 0.062\\
250 & Kendall & 0.375\\
\addlinespace
500 & Pearson & 1.425\\
500 & Spearman & 0.135\\
500 & Kendall & 1.495\\
\bottomrule
\end{tabular}
\end{table}

@@

The results with larger dimensional vectors show scalability to
ultra-high dimensions for all three correlation types, although the
total times do become much larger. Figure @ref(fig:ch040-largeDfig)
displays computation times for \(d=\{1000, 2500, 5000, 10000\}\). For
\(d\) equal to 1000 and 2500, the total time is under a couple of
minutes. At \(d\) of 5000 and 10,000, Pearson correlation matching in
the \emph{Adjust Correlation} step becomes costly. Interestingly,
Pearson is actually faster than Kendall for \(d=10,000\) due to
bottlenecks in \emph{Compute Correlation} and \emph{Check
Admissibility}. Uniformly, matching Spearman correlations is faster,
with total times under 5 minutes for \(d=10,000\), making Spearman
correlations the most computationally-friendly dependency type. With
this in mind, we scaled the simulation to \(d=20,000\) for the Spearman
type and obtained the 1,000 vectors in under an hour (data not shown).
In principle, this would enable the simulation of an entire
human-derived RNA-seq data set. We note that for a given target
correlation matrix and margins, steps i, 1, and 2 only need to be
computed once and the third step, \emph{Simulate Data}, is fast under
all schemes for all \(d\) under consideration.

@@

\caption[Computation times as d increases]{Computation times as d increases.}\label{fig:ch040-largeDfig}
\end{figure}
\end{CodeChunk}

@@

In the bivariate studies, we chose arbitrary simulation parameters for
three distributions, moving from the Gaussian to the discrete and
non-normal, multivariate negative binomial. Under these conditions, the
simulated random vectors sample the desired bivariate distribution
across the entire range of pairwise correlations for the three
dependency measures. The simulation results could differ for other
choices of simulation settings. Specifying extreme correlations near the
boundary or Frechet bounds could result in poor simulation performance.
Fortunately, it is straightforward to evaluate simulation performance by
using strategies similar to those completed above. We expect our random
vector generation to perform well for the vast majority of
NORTA-feasible correlation matrices, but advise to check the performance
before making inferences/further analyses.

Somewhat surprising, Kendall estimation and nearest PD computation scale
poorly compared to Spearman and, even, approximate Pearson matching. In
our experience, Kendall computation times are sensitive to the number of
cores, benefiting from multi-core parallelization. This could mitigate
some of the current algorithmic/implementation shortcomings. Despite
this, Kendall matching is still feasible for most high-dimensional data
sets. Finally, we note that one could use our single-pass algorithm
\texttt{cor\_fastPD} to produce a `close' (not nearest PD) to scale to
even higher dimensions with greater loss of accuracy.

@@

This section demonstrates how to simulate multivariate data using
\texttt{bigsimr}, aiming to replicate the structure of high-dimensional
dependent count data. In an illustration of our proposed methodology, we
seek to simulate RNA-sequencing data by producing simulated random
vectors mimicking the observed data and its generating process. Modeling
RNA-seq using multivariate probability distributions is natural as
inter-gene correlation is an inherent part of biological processes
\citep{Wang2009b}. And yet, many models do not account for this, leading
to major disruptions to the operating characteristics of statistical
estimation, testing, and prediction. See \citet{Efron2012} for a
detailed discussion with related methods and \citet{Wu2012b},
\citet{Schissler2018}; \citet{Schissler2019} for applied examples. The
following subsections apply \texttt{bigsimr}'s methods to real RNA-seq
data, including replicating an estimated parametric structure, MC
probability estimation, and MC evaluation of correlation estimation
efficiency.

@@

We begin by estimating the structure of the TCGA BRCA RNA-seq data set
(see \href{background}{Background}). Ultimately, we will simulate
\(B=10,000\) random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with
\(d=1000\). We assume a multivariate negative binomial (MVNB) model as
RNA-seq counts are often over-dispersed and correlated. Since all \(d\)
selected genes exhibit over-dispersion (data not shown), we proceed to
estimate the NB parameters \((r_i, p_i), i=1,\ldots,d\), to determine
the target marginal PMFs \(f_i\). To complete specification of the
simulation algorithm inputs, we estimate the Spearman correlation matrix
\({ \bf R}_{S}\) to characterize dependency.

@@

Next, we estimate the marginal parameters. We use the method of moments
(MoM) to estimate the marginal parameters for the multivariate negative
binomial model. While, marginal distributions are from the same
probability family (NB), they are heterogeneous in terms of the
parameters probability and size \((p_i, n_i)\) for \(i,\ldots,d\). The
functions below support this estimation for later use in \texttt{rvec}.

@@

We apply these estimators to all 1000 genes across the 878 patients:

@@

Notably, the estimated marginal NB probabilities \(\{ \hat{p}_i \}\) are
small --- ranging in the interval
\([\ensuremath{7.5305\times 10^{-7}} , \ensuremath{5.6592\times 10^{-4}}]\).
This gives rise to highly variable counts and, typically, less
restriction on potential pairwise correlation pairs. Once the functions
are defined/executed to complete marginal estimation, we specify targets
and generate the desired random vectors using \texttt{rvec}. Now we
check admissibility of the specified correlation matrix.

@@

While the exact \(d \times d\) Spearman correlation matrix is not
strictly admissible in our scheme (as seen by the non-positive definite
result above), the approximation is close with a maximum absolute error
of 0.0069 and average absolute error of
\ensuremath{5.3366\times 10^{-4}} across the
\ensuremath{4.995\times 10^{5}} correlations.

\begin{CodeChunk}
\begin{CodeInput}
R> sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
R> colnames(sim_nbinom) <- colnames(brca1000)
\end{CodeInput}
\end{CodeChunk}

Figure @ref(fig:ch050-simDataFig) displays the simulated counts and
pairwise relationships for our example genes from Table
@ref(tab:ch010-realDataTab). Simulated counts roughly mimic the observed
data but with a smoother appearance due to the assumed parametric form
and with less extreme points then the observed data in Figure
@ref(fig:ch010-realDataFig). Figure @ref(fig:ch050-figBRCA) displays the
aggregated results of our simulation by comparing the specified target
parameter (horizontal axes) with the corresponding quantities estimated
from the simulated data (vertical axes). The evaluation shows that the
simulated counts approximately match the target parameters and exhibit
the full range of estimated correlation from the data.

@@

Thus, the possible range of correlation becomes \((-0.881, 0.881)\).
This approximately agrees with the MC results provided.

>>>>

This section describes a LD random vector simulation workflow via the
\texttt{bigsimr} \texttt{R} package. Our \texttt{R} package provides an
interface to the native code written in Julia (registered as the
\texttt{Bigsimr} Julia package. In addition to the native Julia
\texttt{Bigsimr} package and \texttt{R} interface \texttt{bigsimr}, we
also provide a python interface \texttt{bigsimr} that interfaces with
the Julia \texttt{Bigsimr} package. The Julia package provides a
high-performance implementation of our proposed random vector generation
algorithm and associated functions (see Section @ref(algorithms)).

@@

\hypertarget{bivariate-example}{%
\subsection{Bivariate example}\label{bivariate-example}}

We illustrate the use of \texttt{bigsimr} using the New York air quality
data set (\texttt{airquality}) included in the R \texttt{datasets}
package. First, we load the \texttt{bigsimr} library and a few other
convenient data science packages, including the syntactically-elegant
\texttt{tidyverse} suite of \texttt{R} packages. The code chunk below
prepares the computing environment:

@@

Below, we describe a minimal working example --- a bivariate simulation
of two airquality variables: \texttt{Temperature}, in degrees
Fahrenheit, and \texttt{Ozone} level, in parts per billion.

@@

Figure @ref(fig:ch030-aq-joint-dist) visualizes the bivariate
relationship between \texttt{Ozone} and \texttt{Temperature}. We aim to
simulate random two-component vectors mimicking this structure. The
margins are not normally distributed; \texttt{Ozone} level exhibits a
strong positive skew.

@@

\caption[Bivariate scatterplot of Ozone versus Temp with estimated marginal densities]{Bivariate scatterplot of Ozone versus Temp with estimated marginal densities. The Ozone data are modeled marginally as log-normal and the Temperature data as normal.}\label{fig:ch030-aq-joint-dist}
\end{figure}
\end{CodeChunk}

Next, we specify the marginal distributions and correlation coefficient
(both type and magnitude). Here the analyst is free to be creative. For
this example, we avoid goodness-of-fit considerations to determine the
marginal distributions. It is sensible without domain knowledge to
estimate these quantities from the data, and \texttt{bigsimr} contains
fast functions designed for this task.

@@

Based on the estimated densities in Figure
@ref(fig:ch030-aq-joint-dist), we assume \texttt{Temp} is normally
distributed and \texttt{Ozone} is log-normally distributed, as the
latter values are positive and skewed. We use the classical unbiased
estimators for the normal distribution's parameters and maximum
likelihood estimators for the log-normal parameters:

@@

Now, we configure the input marginals for later input into
\texttt{rvec}. The marginal distributions are specified using
\texttt{Julia}'s \texttt{Distributions} package and stored in a vector.

@@

As mentioned, the user must decide how to describe correlation based on
the particulars of the problem. For non-normal data and for improved
simulation accuracy/scalability in our scheme, we advocate the use of
Spearman's \(\rho\) correlation matrix \(R_S\) and Kendall's \(\tau\)
correlation matrix \(R_K\). We also support Pearson correlation
coefficient matching, while cautioning the user to check the performance
for the distribution at hand (see \protect\hyperlink{simulations}{Monte
Carlo evaluations} below for evaluation strategies and guidance). Note
that hese estimation methods are classical approaches, not specifically
designed for high-dimensional correlation estimation.

@@

Once a target correlation matrix is specified, first convert the
dependency type to Pearson to correctly specify the MVN inputs. Then,
check that the converted matrix \(R_X\) is a valid correlation matrix
(PD and 1s on the diagonal). For this bivariate example, \(R_X\) is a
valid correlation matrix. Typically in HD the resultant MVN Pearson
correlation matrix is indefinite and requires approximation (see the HD
examples in subsequent sections).

\begin{CodeChunk}
\begin{CodeInput}
R> # Step 1. Mapping
R> (R_X <- bs$cor_convert(R_S, bs$Spearman, bs$Pearson))
\end{CodeInput}
\begin{CodeOutput}
       [,1]   [,2]
[1,] 1.0000 0.7886
[2,] 0.7886 1.0000
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> # Step 2. Check admissibility
R> bs$iscorrelation(R_X)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

Despite being a valid correlation matrix under this definition, marginal
distributions induce Frechet bounds on the possible Pearson correlation
values. To check that the converted correlation values fall within these
bounds, use \texttt{bigsimr::cor\_bounds} to estimate the pairwise lower
and upper correlation bounds. \texttt{cor\_bounds} uses the Generate,
Sort, and Correlate algorithm of \citet{DH2011}. For the assumed
marginals, the Pearson correlation coefficient is not free to vary in
{[}-1,1{]}, as seen below. Our MC estimate of the bounds slightly
underestimates the theoretic bounds of \((-0.881, 0.881)\). (See an
analytic derivation presented in the Appendix). Since our single Pearson
correlation coefficient is within the theoretical bounds, the
correlation matrix is valid for our simulation strategy.

@@

@@

A single \texttt{rvec} execution simulates the desired \(10,000\) random
vectors from the assumed bivariate distribution of \texttt{Ozone} and
\texttt{Temp}. Note that the target input in specified using the
pre-computed MVN correlation matrix.

@@

In this section, we conduct MC studies to investigate method
performance. Marginal parameter matching is straightforward in our
scheme as it is a sequence of univariate inverse probability transforms.
This process is both fast and exact. On the other hand, accurate and
computationally efficent dependency matching presents a challenge,
especially in the HD setting. To evaluate our methods in those respects,
we design the following numerical experiments to first assess accuracy
of matching dependency parameters in bivariate simulations and then time
the procedure in increasingly large dimension \(d\).

@@

We select bivariate simulation configurations to ultimately simulate our
motivating discrete-valued RNA-seq example. Thus, we proceed in
increasing departure from normality, leading to the multivariate
negative binomial (MVNB) model for our motivating data. We begin with
empirically evaluating the dependency matching across all three
supported correlations --- Pearson, Spearman, and Kendall --- in
identical, bivariate marginal configurations. For each pair of identical
margins, we vary the target correlation across \(\Omega\), the set of
possible admissible values for each correlation type, to evaluate the
simulation's ability to simulate all possible correlations. The
simulations progress from bivariate normal, to bivariate gamma
(non-normal yet continuous), and bivariate negative binomial (non-normal
and discrete).

Table @ref(tab:sims) lists our identical-marginal, bivariate simulation
configurations. We increase the simulate replicates \(B\) to visually
ensure that our results converge to the target correlations and gauge
efficiency. We select distributions beginning with a standard
multivariate normal (MVN) as we expect the performance to be exact (up
to MC error) for all correlation types. Then, we select a non-symmetric
continuous distribution: a standard (rate =1), two-component
multivariate gamma (MVG). Finally, we select distributions and marginal
parameter values that are motivated by our RNA-seq data, namely values
proximal to probabilities and sizes estimated from the data (see
\href{examples}{RNA-seq data application} for estimation details).
Specifically, we assume a MVNB with
\(p_1 = p_2 = 3\times10^{-4}, r_1 = r_2 = 4, \rho \in \Omega\).

For each of the unique 9 simulation configurations described above, we
estimate the correlation bounds and vary the correlations along a
sequence of 100 points evenly placed within the bounds, aiming to
explore \(\Omega\). Specifically, we set correlations
\(\{ \rho_1 = ( \hat{l} + \epsilon), \rho_2 = (\hat{l} + \epsilon) + \delta, \ldots, \rho_{100} = (\hat{u} - \epsilon) \}\),
with \(\hat{l}\) and \(\hat{u}\) being the estimated lower and upper
bounds, respectively, with increment value \(\delta\). The adjustment
factor, \(\epsilon=0.01\), is introduced to handle numeric issues when
the bound is specified exactly.

@@

Overall, the studies show that our methodology is generally accurate
across the entire range of possible correlation for all three dependency
measures. Our Pearson matching performs nearly as well as Spearman or
Kendall, except for a slight increase in error for negative binomial
case.

@@

With information of our method's accuracy from a low-dimensional
perspective, we now assess whether \texttt{bigsimr} can scale to larger
dimensional problems with practical computation times. We ultimately
generate \(B=1,000\) random vectors for
\(d=\{100, 250, 500, 1000, 2500, 5000, 10000\}\) for each correlation
type, \{Pearson, Spearman, Kendall\} while timing the algorithm's major
steps.

We first produce a synthetic ``data set'' by completing the following
steps:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Produce heterogeneous gamma marginals by randomly selecting the
  \(j^{th}\) gamma shape parameter from
  \(U_j \sim uniform(1,10), j=1,\ldots,d\) and the \(j^{th}\) rate
  parameter from \(V_j \sim exp(1/5), j=1,\ldots,d\), with the constant
  parameters determined arbitrarily.
\item
  Produce a random full-rank Pearson correlation matrix via
  \texttt{cor\_randPD} of size \(d \times d\).
\item
  Simulate a ``data set'' of \(1,000 \times d\) random vectors via
  \texttt{rvec}. \setstretch{2.0}
\end{enumerate}

With the synthetic data set in hand, we complete and time the following
four steps involved in a typical workflow (including correlation
estimation).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Estimate the correlation matrix from the ``data'' in the \emph{Compute
  Correlation} step.
\item
  Map the correlations to initialize the algorithm (Pearson to Pearson,
  Spearman to Pearson, or Kendall to Pearson) in the \emph{Adjust
  Correlation} step.
\item
  Check whether the mapping produces a valid correlation matrix and, if
  not, find the nearest PD correlation matrix in the \emph{Check
  Admissibility} step.
\item
  Simulate \(1,000\) vectors in the \emph{Simulate Data} step.
  \setstretch{2.0}
\end{enumerate}

The experiments are conducted on a MacBook Pro carrying a 2.4 GHz 8-Core
Intel Core i9 processor, with all 16 threads employed during
computation. Table @ref(tab:ch040-moderateDtab) displays the total
computation time for moderate dimesions (\(d \leq 500\)). In every
simullation setting, the 1,000 vectors are generated rapidly, executing
in under 2 seconds.

@@

\caption{\label{tab:tab:ch040-moderateDtab}Total time to produce 1,000 random vectors with a random correlation matrix and hetereogeneous gamma margins.}
\centering
\begin{tabular}[t]{llr}
\toprule
Dimension & Correlation type & Total Time (Seconds)\\
\midrule
100 & Pearson & 0.081\\
100 & Spearman & 0.027\\
100 & Kendall & 0.075\\
\addlinespace
250 & Pearson & 0.378\\
250 & Spearman & 0.062\\
250 & Kendall & 0.375\\
\addlinespace
500 & Pearson & 1.425\\
500 & Spearman & 0.135\\
500 & Kendall & 1.495\\
\bottomrule
\end{tabular}
\end{table}

@@

The results for \(d > 500\) show scalability to ultra-high dimensions
for all three correlation types, although the total times do become much
larger. Figure @ref(fig:ch040-largeDfig) displays computation times for
\(d=\{1000, 2500, 5000, 10000\}\). For \(d\) equal to 1000 and 2500, the
total time is under a couple of minutes. At \(d\) of 5000 and 10,000,
Pearson correlation matching in the \emph{Adjust Correlation} step
becomes costly. Interestingly, Pearson is actually faster than Kendall
for \(d=10,000\) due to bottlenecks in \emph{Compute Correlation} and
\emph{Check Admissibility}. Uniformly, matching Spearman correlations is
faster, with total times under 5 minutes for \(d=10,000\), making
Spearman the most computationally-friendly dependency type for our
package. With this in mind, we scaled the simulation to \(d=20,000\) for
the Spearman type and obtained the 1,000 vectors in under an hour (data
not shown). In principle, this would enable the simulation of an entire
human-derived RNA-seq data set. We note that for a given target
correlation matrix and margins, steps 1, 2, and 3 only need to be
computed once and the fourth step, \emph{Simulate Data}, is nearly
instaneous for all settings considered.

@@

\caption[Computation times for HD multivariate gamma simulation]{Computation times for HD multivariate gamma simulation.}\label{fig:ch040-largeDfig}
\end{figure}
\end{CodeChunk}

@@

In the bivariate studies, we chose arbitrary simulation parameters for
three distributions, moving from the Gaussian to discrete and non-normal
MVNB. Under these conditions, the simulated random vectors sample the
desired bivariate distribution across the entire range of pairwise
correlations for the three dependency measures. The simulation results
could differ for other choices of simulation settings. Specifying
extreme correlations near the boundary or Frechet bounds could result in
poor simulation performance. Fortunately, it is straightforward to
evaluate simulation performance by using strategies similar to those
completed above. We expect our random vector generation to perform well
for the vast majority of NORTA-feasible correlation matrices, but advise
to check the performance before making inferences/further analyses.

Somewhat surprising, Kendall estimation and nearest PD computation scale
poorly compared to Spearman and, even, approximate Pearson matching. In
our experience, Kendall computation times are sensitive to the number of
cores, benefiting from multi-core parallelization. This could mitigate
some of the current algorithmic/implementation shortcomings. Despite
this, Kendall matching is still feasible for most HD data sets. Finally,
we note that one could use our single-pass algorithm
\texttt{cor\_fastPD} to produce a `close' (not nearest PD) to scale to
even higher dimensions with some loss of accuracy.

@@

This section demonstrates how to simulate multivariate data using
\texttt{bigsimr}, aiming to replicate the structure of HD dependent
count data. In an illustration of our proposed methodology, we seek to
simulate RNA-sequencing data by producing simulated random vectors
mimicking the observed data and its generating process. Modeling RNA-seq
using multivariate probability distributions is natural as inter-gene
correlation occurs during biological processes \citep{Wang2009b}. And
yet, many models do not account for this, leading to major disruptions
to the operating characteristics of statistical estimation, testing, and
prediction. See \citet{Efron2012} for a detailed discussion with related
methods and \citet{Wu2012b}, \citet{Schissler2018};
\citet{Schissler2019} for applied examples. The following subsections
apply \texttt{bigsimr}'s methods to real RNA-seq data, including
replicating an estimated parametric structure, probability estimation,
and evaluation of correlation estimation efficiency.

@@

We begin by estimating the structure of the TCGA BRCA RNA-seq data set.
Ultimately, we will simulate \(B=10,000\) random vectors
\({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \(d=1000\). We assume a MVNB
model as RNA-seq counts are often over-dispersed and correlated. Since
all \(d\) selected genes exhibit over-dispersion (data not shown), we
proceed to estimate the NB parameters \((r_i, p_i), i=1,\ldots,d\), to
determine the target marginal PMFs \(f_i\). To complete specification of
the simulation algorithm inputs, we estimate the Spearman correlation
matrix \(R_S\) to characterize dependency.

@@

Next, we estimate the marginal parameters. We use method of moments to
estimate the marginal parameters for the multivariate negative binomial
model. The marginal distributions are from the same probability family
(NB), yet they are heterogeneous in terms of the parameters probability
and size \((p_i, n_i)\) for \(i,\ldots,d\). The functions below support
this estimation for later use in \texttt{rvec}.

@@

We apply these estimators to the highest-expressing 1000 genes across
the 878 patients:

@@

Notably, the estimated marginal NB probabilities \(\{ \hat{p}_i \}\) are
small --- ranging in the interval
\([\ensuremath{7.5305\times 10^{-7}} , \ensuremath{5.6592\times 10^{-4}}]\).
This gives rise to highly variable counts and, thus, less restriction on
potential pairwise correlation pairs. Given the marginals, we now
specify targets and check admissibility of the specified correlation
matrix.

@@

As seen above \(R_X\) is not strictly admissible in our scheme (as seen
by the non-positive definite result above). However, the approximation
is close with a maximum absolute error of 0.0069 and average absolute
error of \ensuremath{5.3366\times 10^{-4}} across the
\ensuremath{4.995\times 10^{5}} correlations. With the inputs
configured, \texttt{rvec} is executed to produce a synthetic RNA-seq
data set:

\begin{CodeChunk}
\begin{CodeInput}
R> sim_nbinom <- bs$rvec(10000, R_X, nb_margins) 
\end{CodeInput}
\end{CodeChunk}

Figure @ref(fig:ch050-simDataFig) displays the simulated counts and
pairwise relationships for our example genes from Table
@ref(tab:ch010-realDataTab). Simulated counts roughly mimic the observed
data but with a smoother appearance due to the assumed parametric form
and with less extreme points then the observed data in Figure
@ref(fig:ch010-realDataFig). Figure @ref(fig:ch050-figBRCA) compares the
specified target parameter (horizontal axis) with the corresponding
quantities estimated from the simulated data (vertical axis). The
evaluation shows that the simulated counts approximately match the
target parameters and exhibit the full range of estimated correlations
from the data.

@@

Thus, the possible range of correlation becomes \((-0.881, 0.881)\).
This agrees with the MC results provided.
